---
title: "Week 4 Exercises"
author: "Daniel Jackson"
date: "July 29, 2023"
output: pdf_document
---

Please complete all exercises below. You may use any library that we have covered in class. The data we will be using comes from the tidyr package, so you must use that. 

1) Examine the who and population data sets that come with the tidyr library. the who data is not tidy, you will need to reshape the new_sp_m014 to newrel_f65 columns to long format retaining country, iso2, iso3, and year. The data in the columns you are reshaping contains patterns described in the details section below. You will need to assign three columns: diagnosis, gender, and age to the patterns described in the details. 

Your tidy data should look like the following:
  country     iso2  iso3   year diagnosis gender age   count
  <chr>       <chr> <chr> <int> <chr>     <chr>  <chr> <int>
1 Afghanistan AF    AFG    1980 sp        m      014      NA
2 Afghanistan AF    AFG    1980 sp        m      1524     NA
3 Afghanistan AF    AFG    1980 sp        m      2534     NA
4 Afghanistan AF    AFG    1980 sp        m      3544     NA
5 Afghanistan AF    AFG    1980 sp        m      4554     NA
6 Afghanistan AF    AFG    1980 sp        m      5564     NA

Details
The data uses the original codes given by the World Health Organization. The column names for columns five through 60 are made by combining new_ to a code for method of diagnosis (rel = relapse, sn = negative pulmonary smear, sp = positive pulmonary smear, ep = extrapulmonary) to a code for gender (f = female, m = male) to a code for age group (014 = 0-14 yrs of age, 1524 = 15-24 years of age, 2534 = 25 to 34 years of age, 3544 = 35 to 44 years of age, 4554 = 45 to 54 years of age, 5564 = 55 to 64 years of age, 65 = 65 years of age or older).

*Note: use data(who) and data(population) to load the data into your environment.*
*Use the arguments cols, names_to, names_pattern, and values_to.*
*Your regex should be = ("new_?(.*)_(.)(.*)")*

https://tidyr.tidyverse.org/reference/who.html

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Read in the who and population data frames
data(who)
data(population)

# Use pivot_longer() function. Use cols from 
who = who %>%
  pivot_longer(cols = new_sp_m014:newrel_f65,
               names_to = c("diagnosis", "gender", "age"),
               names_pattern = "new_?(.*)_(.)(.*)",
               values_to = "count")
head(who)

```

2) There are two common keys between the data sets, with who as the left table, join the population data by country and year so that the population is available within the who dataset. 


```{r}
# Use left_join() to join the keys from population to keys in who data set.
## Use join_by() argument to join by different variables: country and year.
who_pop_df = who %>%
  left_join(population, join_by(country,year))
head(who_pop_df)

```

3) Split the age column into two columns, min age and max age. Notice that there is no character separator. Check the documentation with ?separate to understand other ways to separate the age column. Keep in mind that 0 to 14 is coded as 014 (3 characters) and the other age groups are coded with 4 characters. 65 only has two characters, but we will ignore that until the next problem. 

```{r}
# Separate age column into two columns: min_age and max_age. Used separator()
## function and used sep = -2 which sep is intepreted as character positions to split
## at. -2 would split after the second to last integer in numeric value
who_pop_df = who_pop_df %>%
  separate(age, 
           sep = -2,
           into = c('min_age', 'max_age'))
head(who_pop_df)

```

4) Since we ignored the 65+ group in the previous problem we will fix it here. If you examine the data you will notice that 65 was placed into the max_age column and there is no value for min_age for those records. To fix this use mutate() in order to replace the blank value in the min_age column with the value from the max_age column and another mutate to replace the 65 in the max column with an Inf. Be sure to keep the variables as character vectors.  

```{r}
# Use mutate() function to change values within data frame. Used two ifelse() functions
## to see where min_age is blank to replace it with max_age, and to see where
### max_age is equal to 65, to replace it with "Inf".
who_pop_df = who_pop_df %>%
  mutate(min_age = ifelse(min_age == "",
                          max_age,
                          min_age)) %>%
  mutate(max_age = ifelse(max_age == "65",
                          "Inf",
                          max_age))
head(who_pop_df)
```

5) Find the count per diagnosis for males and females.

*See ?sum for a hint on resolving NA values.*

```{r}
# Want to group by gender and diagnosis, then summarize by the sum of count
## column. Using na.rm = TRUE is how we solve dealing with the NA values
count_per_gender_diag = who_pop_df %>%
  group_by(gender, diagnosis) %>%
  summarize(count = sum(count, na.rm = TRUE))
head(count_per_gender_diag)

```

6) Now create a plot using ggplot and geom_col where your x axis is gender, your y axis represents the counts, and facet by diagnosis. Be sure to give your plot a title and resolve the axis labels. 

```{r}
# Using ggplot() and geom() point. Make x axis as gender, y axis as count and
## use facet_grid() to facet by diagnosis. Use labs() to name axes and give a 
### title.
ggplot(count_per_gender_diag) +
  geom_col(aes(x = gender, 
               y = count, 
               fill = gender)) +
    facet_grid(.~diagnosis) +
    labs(x = 'Gender', y = 'Counts', title = 'Gender Counts by Diagnosis')
```

7) Find the percentage of population by year, gender, and diagnosis. Be sure to remove rows containing NA values. 

```{r}
# Group by year, gender and diagnosis. Add a column called pop_percent that equals
## to the sum of count divided by the average population. Use na.rm = TRUE to deal
### with NA values. Then use drop_na() to drop any NA valued rows.
population_percentage = who_pop_df %>%
  group_by(year, gender, diagnosis) %>%
  summarize(pop_percent = sum(count,na.rm = TRUE)/mean(population, na.rm = TRUE))%>%
  drop_na()
head(population_percentage)
```

8) Create a line plot in ggplot where your x axis contains the year and y axis contains the percent of world population. Facet this plot by diagnosis with each plot stacked vertically. You should have a line for each gender within each facet. Be sure to format your y axis and give your plot a title. 

```{r}
# Use ggplot() and geom_line() and make x axis year, y axis pop_percent, and 
## color to gender. Facet by diagnosis then label each axis and give a title. Used
#### scale_y_continuous() to add percents to y axis numbers.
population_percentage %>%
ggplot(aes(x = year,
           y = pop_percent,
           color = gender)) +
      geom_line() +
         facet_grid(.~diagnosis) +
         labs(x = 'Year', 
              y = 'Percent of World Population',
              title = 'Percent of World Population by Year, Gender. and Diagnosis') +
      scale_y_continuous(labels = scales::percent_format(scale = 1))
```

9) Now unite the min and max age variables into a new variable named age_range. Use a '-' as the separator. 

```{r}
# Use unite() function and name new column 'age_range' and combine min_age and 
## max_age in arguments with sep = '-'
who_pop_df = who_pop_df %>%
  unite(col = 'age_range', min_age : max_age, sep = '-')
head(who_pop_df$age_range)

```

10) Find the percentage contribution of each age group by diagnosis. You will first need to find the count of all diagnoses then find the count of all diagnoses by age group. Join the former to the later and calculate the percent of each age group. Plot these as a geom_col where the x axis is the diagnosis, y axis is the percent of total, and faceted by age group. 

```{r}
# Create data frame that counts all diagnoses
diagnosis_count = who_pop_df %>%
  group_by(diagnosis) %>%
  summarize(total_diag_count = sum(count, na.rm = TRUE))

# Create data frame counting all diagnoses by age group
diagnosis_count_by = who_pop_df %>%
  group_by(diagnosis, age_range) %>%
  summarize(total_diag_count_age = sum(count, na.rm = TRUE))

# Join the two data frames by common key, which is diagnosis
join_diagnosis = diagnosis_count %>%
  left_join(diagnosis_count_by, by = 'diagnosis')

# Do a new data frame, group by diagnosis and age_range then add a new column that
## takes diagnosis count by age and divids by total diagnosis count and multiply
### by 100 to get percentage.
diagnosis_percentage_by_age_group = join_diagnosis %>%
  group_by(diagnosis, age_range) %>%
  summarize(diag_percent = (total_diag_count_age / total_diag_count) * 100)
head(diagnosis_percentage_by_age_group)

# Create geom_col() plot with x as diagnosis, y as diagnosis, faceted by age group.
## Label x and y axis, and give title. Add percentages to y axis.
diagnosis_percentage_by_age_group %>%
  ggplot(aes(x = diagnosis, 
             y = diag_percent,
             fill = diagnosis)) +
  geom_col() +
  facet_grid(.~age_range) +
  labs(x = 'Diagnosis', 
       y = 'Percentage',
       title = 'Percentage of Age Group by Diagnosis') +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

```