---
title: "Week 2 Exercises"
author: "Daniel Jackson"
date: "July 15, 2023"
output: pdf_document
---

Please complete all exercises below. You may use stringr, lubridate, or the forcats library. 

Place this at the top of your script:
library(stringr)
library(lubridate)
library(forcats)

# Exercise 1
Read the sales_pipe.txt file into an R data frame as sales. 

```{r}
library(stringr)
library(lubridate)
library(forcats)

# Use guess_encoding() from readr library to find out that the encoding in the 
## file is ISO-8859-1.
readr::guess_encoding("Data/sales_pipe.txt") 

# Use the argument fileEncoding = 'ISO-8859-1' to properly read in sales_pipe.txt. 
## Used sep = "|" argument since values in file are separated by "|" character.
sales = read.delim("Data/sales_pipe.txt", 
                   fileEncoding="ISO-8859-1",
                   stringsAsFactors=FALSE,
                   sep = "|") 

```

# Exercise 2
You can extract a vector of columns names from a data frame using the colnames() function. Notice the first column has some odd characters. Change the column name for the FIRST column in the sales date frame to Row.ID. 

**Note: You will need to assign the first element of colnames to a single character.**

```{r}
# Using str_replace() and colnames(), we can replace anywhere it says ï..Row.ID"
## to "Row.ID" in the columns within sales dataframe. Assign it to variable 
### sales_columns
sales_columns  = str_replace(colnames(sales),"ï..Row.ID","Row.ID")

# Assign sales_columns variable we just created to colnames(sales) to update 
## dataframe column names
colnames(sales) = sales_columns
print(sales_columns)

```

# Exercise 3
Convert both Order.Date and Ship.Date to date vectors within the sales data frame. 

What is the number of days between the most recent order and the oldest order?
  Time difference is 1457 days.
How many years is that?
  3.99 years
How many weeks?
  207.57 weeks

**Note: Use lubridate**

```{r}
# Shows structure of elements in Order.Date vector. Elements are characters.
str(sales$Order.Date) 

# Use as.Date() function to convert elements in Order.Date vector from 
## characters to dates.
sales$Order.Date = as.Date(sales$Order.Date,format = '%m/%d/%Y') 

# Use inherits() function to verify elements in Order.Date are now dates. 
## This returned TRUE.
inherits(sales$Order.Date, c("Date")) 

# Shows elements in Ship.Date vector are characters. Need to convert elements
## from characters to dates.
str(sales$Ship.Date)

# Convert elements in Ship.Date to dates
sales$Ship.Date = as.Date(sales$Ship.Date,format = '%B %d %Y')

# Confirm elements are dates with inherits() function.
inherits(sales$Ship.Date, c("Date")) 

# Assign newest_order to max (newest) date of Order.Date
newest_order = max(sales$Order.Date) 

# Assign oldest_order to min (oldest) date of Order.Date
oldest_order = min(sales$Order.Date)

# The newest order date is represented by the max, which returns "2017-12-30". 
## The oldest order date is represented by the min, which is "2014-01-07" 
### When you subtract the two, it returns a time difference of 1457 days.
order_difference = (newest_order) - (oldest_order) 

# To get time difference between orders in years (assigned to order_difference_years), 
## we divide order_difference by 365 (365 days in a year) to get 3.99 years.
order_difference_years = (order_difference)/365 

# To get time difference between orders in weeks (assigned to order_difference_weeks),
## we multiply order_difference_years by 52 (52 weeks in a year) to get 207.57 weeks.
order_difference_weeks = (order_difference_years)*52

```


# Exercise 4
What is the average number of days it takes to ship an order?

```{r}
# Take the difference between Ship.Date and Order.Date and found the average of
## those days. Assigned this value to avg_days_to_ship. This returned 3.91 days.

avg_days_to_ship = mean(sales$Ship.Date-sales$Order.Date)
```


# Exercise 5
How many customers have the first name Bill?
You will need to split the customer name into first and last name segments and then use a regular expression to match the first name bill. Use the length() function to determine the number of customers with the first name Bill in the sales data. 

```{r}
# Create character vector that removes any duplicate customers in Customer.Name column
unique_customer_name_vec = unique(sales$Customer.Name)

# Create a dataframe that splits Customer.Name where first and last name are 
## each of the two columns.
split_customer_name_df = stringr::str_split_fixed(string=unique_customer_name_vec, 
                                               pattern = ' ',
                                               n=2)

# Using length() function, we subset our data frame to see how many times the 
## name Bill shows up and assign it to num_cust_named_bill. This returns a length
### of 6. So, 6 customers are named Bill.
num_cust_named_bill = length(split_customer_name_df
                             [split_customer_name_df == "Bill"])
```

# Exercise 6
How many mentions of the word 'table' are there in the Product.Name column? 
**Note you can do this in one line of code**

```{r}
# Use str_count() to count number of times 'table' appears in Product.Name column. 
## Use sum() function to add those together to get 240 and assign it to 
### mentions_of_table_val value. 240 mentiuons of 'table' in Product.Name column
mentions_of_table_val = sum(str_count(sales$Product.Name, 'table'))
```

# Exercise 7
Create a table of counts for each state in the sales data. The counts table should be ordered alphabetically from A to Z. 

```{r}
# Use table() function to create table of counts for each state. The table 
## defaults to ordering states alphabetically from A to Z. Assign table to sales_table.
sales_table = table(sales$State)

# Print sales_table
print(sales_table)
```

# Exercise 8
Create an alphabetically ordered barplot for each sales Category in the State of Texas. 

```{r}
# Create Texas only data frame table by subsetting sales data frame to only 
## include rows that include Texas as a state, and using the sales_columns 
### vectors we defined earlier as our columns.
texas_df = sales[sales$State == 'Texas',  c(sales_columns)]

# Factor the category column within the new dataframe
texas_df$Category = factor(texas_df$Category)

# Use the table function on the factored column. This will alphabetize the 
## levels within the factor
texas_category_table = table(texas_df$Category)

# Print barplot  of that table
texas_category_barplot = barplot(texas_category_table)
```

# Exercise 9
Find the average profit by region.
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r}
# Using aggregate() function, the "~" argument takes the Profit numbers per Region, 
## then uses "FUN = mean" to say we want the mean of the Profit per Region. Assign it to new variable. 

avg_profit_region = aggregate(sales$Profit ~ sales$Region, FUN = mean)

# Average profit by Central region is 20.47, by East region is 29.92, by South 
## is 11.28 and by West is 32.77
```

# Exercise 10
Find the average profit by order year. 
**Note: You will need to use the aggregate() function to do this. To understand how the function works type ?aggregate in the console.**

```{r}
# Need to split Order.date column 
order_date_split = str_split_fixed(sales$Order.Date, 
                                   pattern = "-", n = 3)

# Use aggregate() function to find average sales profit by year. order_date_year 
## is first column of order_date_split
order_date_year = order_date_split[,1]
avg_profit_year = aggregate(sales$Profit ~ order_date_year, 
                            FUN = mean)
# Average profit by year for 2014 is 32.25, for 2015 is 21.59, for 2016 30.11 
## and for 2017 is 21.32
```